Deep mutational scanning of hemagglutinin helps predict evolutionary fates of human H3N2 influenza variants

### 2018-06-23 paper/reviewer_response_PNAS.tex

\response{We have validated that one of these mutations in the transmembrane domain, C199(HA2)K, reaches reverse-genetics titers comparable to wildtype.
These data are in the new Figure S7.
In addition, various other studies have found that chimeric HAs with transmembrane domains from completely different proteins can still maintain HA cell surface expression and fusion activity (Melikyan et al 1999 Mol Biol Cell, Schroth-Diez et al 1998 JVI, Dong et al 1992 JVI, Roth et al 1986 JCB).

We posit that the disparity between the conservation of the transmembrane domain in nature and its apparent mutational tolerance in our experiments could arise at least in part from a lack of immune pressure for mutations in nature, as discussed in the response to the previous point.
}

4) Were sequences from passaged viruses (isolated on MDCKs and other cell lines) included in the comparison between in vitro mutagenesis data and natural evolution data? They need to be excluded since their presence might induce a bias towards what is tolerated/preferred in cell culture. Only sequences obtained directly from patient specimens should be used for the comparison.

\response{
This is an excellent question.
We now perform a separate analysis of the relationship between our measured mutational effects and maximum frequency in nature for mutations from unpassaged isolates only.
Because sequences from unpassaged isolates have only become common relatively recently, we could perform this analysis only for the post-Perth/2009 partition of the H3N2 HA phylogenetic tree.
The new Figure 5 shows that the correlation between our measurements and natural evolution is actually \emph{stronger} when we look only at unpassaged isolates.
This makes sense---lab passage mutations will not do well in nature, but will be scored favorably in our experiments.
Therefore, it makes sense that our approach works \emph{even better} when the analysis is restricted to unpassaged isolates.
We thank the reviewer for this suggestion, which has strengthened our work.
}

@@

We think that the reason that washing away the original inoculum has no discernible effect on the results it that the low concentration of virus in the inoculum makes a negligible contribution to the total number of virions that are present at the time of sequencing, since vastly more new virions are produced from the infected cells.
}

c) This would actually not be an issue, if the dilution from the rescue culture to the new culture would be large enough. But this is not the case. The authors passaged $9\times10^5$ TCID50 (and likely much more DIs/RNA) from the rescue cultures onto the new cells in a volume of 25 ml. This is, depending on experiment, between 10 and 1 ml of supernatant of the original rescue and leads to a 1:2.5 or 1:25 dilution respectively.

\response{
We apologize if the original wording of our methods were unclear. 
For each mutant virus library, we passaged a total of $9\times10^5$ TCID50, but this was passaged across \emph{fifteen} 15-cm dishes \emph{each} of which received 25 ml.
Because we used an inoculum of 2.5 TCID50 / $\mu$l across these fifteen dishes, this corresponds to dilutions that range from 1:37 to 1:294.
The number of dishes used to passage each mutant virus library is now mentioned in the Methods; before it had to be calculated from the stated total number of cells and the cells per dish.

Also, we note that the supernatants from the helper virus rescues are collected at just 24 hours post-infection, which leads to lower viral titers (which is why we have to passage relatively large volumes), but also presumably much lower amounts of DIs / RNA.
This is probably the reason why (as described above) removing the inoculum did not discernibly affect the results.
}

@@

d) The supernatant is never centrifuged before RNA extraction. Cell debris is not removed which is likely contributing significantly to RNA sequences detected in the supernatant. However, RNA associated with cell debris is not representative of RNA from fit, replicating viruses.

\response{We apologize for failing to provide sufficiently detailed methods stating that we ultra-centrifuged \emph{clarified} the viral supernatant before RNA extraction.
The supernatant was clarified by centrifugation at 2000 $\times$ \textit{g} for five minutes to remove any cellular debris.
We have updated the Methods to now include this detail, and (as mentioned above) also now include our full lab notes so that all other details can be reviewed.
Specifically, see \url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript/blob/master/lab_notes/Perth09_HA_Library_Rescue.pdf} (relevant details of the notebook can be found in pages 21-25)
}

@@

The experiments should be repeated with at least one more passaging step, proper removal of inoculum and washing of the cells after infection. Centrifugation steps should also be included between the passages and before RNA extraction in the end.

\response{For the reasons described above, it is not necessary to repeat the experiments.
We do not deny that there are slightly different ways that the experiments could be done, but the in end, the ``proof is in the pudding.'' 
In the manuscript, we show that the measurements made using the experimental parameters that we chose lead to useful data as judged by the following criterion:
\begin{itemize}
\item The results from fully independent experimental replicates are highly correlated (Fig.~1D), which would not be the case if the results were dominated by noise.
\item The experimental measurements can inform substitution models that describe influenza virus's evolution much better than standard models (Table~1).
\item The experimental measurements provide information that can help predict the fates of influenza virus lineages in nature (Fig.~5).
\end{itemize}
For any experimental study there are an almost infinite number of slight methodological variations that could be employed, and it is impossible to test them all.
Our experiments used reasonable parameter choices, were performed in full biological triplicate, and give meaningful / useful results with respect to the larger goals of the study.}

@@

The authors are upfront about the limited degree to which their calculated mutational effect and the maximum mutation frequency in nature correlate, describing it as ``modest.'' One of the major points of the paper is that conclusions from this type of analysis cannot be generalized when using data from strains that are too divergent. The authors should address the possibility that data from deep mutational scanning can be generalized to divergent strains if the study is done in a system that is more reflective of nature.

\response{This is a good point, and we have added the following sentence to the Discussion: ``We suspect that performing similar experiments using more realistic and complex selections (e.g., ferrets or primary human airway cultures) might further improve their utility and possibly their generalizability to more divergent strains."}

The experimental approach of deep mutational scanning has been applied to viral proteins before, including HIV env, influenza H1 HA, and NP. The novelty in this paper is that it is the first time such an analysis has been performed for H3 specifically and that it demonstrates the (albeit limited) utility of deep mutational scanning in predicting the success of viral strains in nature. Ultimately, it is clear from the discussion that the authors see this study as an intermediate step, suggesting that their work is "an improvement over no information at all." While something is indeed better than nothing, in my opinion, this paper represents only an incremental step forward in the field of influenza evolutionary forecasting due to the aforementioned limitations, especially since this experiment and subsequent analysis could be performed using more stringent selection methods during viral passaging like treatment with human sera for data that is presumably more faithful to nature.

\response{The reviewer is correct that additional experiments might lead to further improvements.
However, ours is the first paper to show that data from deep mutational scanning can actually help predict real evolutionary outcomes in nature---which is an important step forward that merits publication.

We are excited about the future directions that the reviewer suggests.
However, the current study is a complete and important contribution even if (as is almost invariably the case in science) future work can probably take things further.
Researchers have been performing functional experiments on influenza virus for decades, and have been trying to forecast the virus's evolution for decades---yet ours is the first study to successfully use such functional experiments to predict the success of viral strains in nature. 
We have gone to great pains to fairly discuss limitations and caveats and how things could be improved in the future, but that should not detract from the advances in the current paper.
}

### 2018-06-23 paper/reviewer_response_PNAS.tex

Of note, position 1 for the H1 HA in the supplementary figure is not even shown.

\response{
The reviewer is correct. 
Figure~S3 shows data for a H1 HA as measured previously by Doud and Bloom (2016), and these prior data are simply presented here in H3 numbering for comparative purposes (the original Doud and Bloom paper used a different HA numbering scheme that makes site-to-site comparisons with the current study difficult). 
The Doud and Bloom study used mutant libraries in which all codons in HA \emph{except} for the one encoding the N-terminal methionine were mutagenized.
Therefore, that study did not estimate the effects of mutations to the first codon, which is why it is not shown in Figure~S3.
We have added a sentence to the legend for this supplementary figure explaining this fact.
}

2) While other important features of the HA like several cysteines seem to be conserved, there is no pressure on other elements which are considered important for viral fitness including the NGT glycosylation site in the beginning of HA1 which is conserved in all group 2 HAs or the adjacent NAT glycosylation site which is conserved in H3, H7, H10 and H15. On the other hand, head glycosylation sites, which are highly variable in nature, seem to be conserved. Again, viruses without these glycosylation sites should be rescued and evaluated for fitness relative to wild type.

\response{This is a valid point, and we have now also individually evaluated the reverse-genetics titers of three glycosylation site mutants.
Two of the mutants, T24F and T40V, remove the aforementioned highly conserved but mutationally tolerant NGT and NAT glycosylation sites near the start of HA1.
The third mutation, S287A, removes a glycosylation site that is highly variable in nature but that our experiments suggest is under functional constraint.
The new Figure S5 shows that the T24F and T40V mutants grow to titers comparable to wildtype, supporting the conclusion that they are expendable at least in cell culture. 
However, the titer of the S287A is modestly lower than wildtype, supporting the conclusion that this glycan contributes to viral growth in cell culture.

@@

One possible explanation is simply that cell culture does not fully capture the constraints in nature.
But a more interesting possibility is that conservation in nature is not expected to strictly parallel mutational tolerance.
The tendency for a site to change in nature depends on how much selection there is for the mutations.
For instance, Wei et al (\textit{Science Translational Medicine}, 2:24ra21, 2010) have pointed out that glycosylation motifs evolved rapidly in the globular head of HA from human H1 influenza in the period from 1918 to 2008 (seasonal H1N1), but changed very little in the same period in the swine H1 lineages descended from the 1918 virus.
The likely reason is differential strengths of selection: human influenza is under strong pressure from accumulated immune memory, whereas swine influenza is under less immune pressure because pigs are short lived and so are often naive when infected with influenza.
This fact highlights that even mutationally tolerant sites will only change rapidly if there is a benefit to doing so, while even somewhat constrained sites can change if there is sufficient pressure.
A strength of our experiments is that we have controlled the selection pressure (selection only for viral growth), and so can measure mutational tolerance in the absence of the confounding effects of immunity.

In addition to the data in Figure S5, we have added the following text:

@@

\response{We individually validated viral growth of a C52A mutant, which removes a Cys critical in forming a disulfide bond with Cys-277 and maintaining HA structure.
We confirmed that this mutant grows very poorly when rescued, and mutations at site 52 never appear in sequenced isolates.

As another example, mutation N170H fails to reach $>$1\% frequency in nature, and according to our dataset, is a mutation towards a highly unpreferred amino acid. 
}

3) The authors picked 31 random clones to evaluate the mutation rate per clone. These sequences should be made available. Did this set include viruses without a proper start codon (see above)? What was their growth phenotype in vitro?

\response{We have now added the list of mutations from the 31 clones to Dataset S3.
Two of the clones had a mutation that removed the ATG start codon.
We did not test the growth of these clones. 
As explained in Figure~S2, these were plasmid variants chosen randomly to Sanger sequence to evaluate the mutation rate prior to creating virus libraries; they were not evaluated as viruses.
The new Figures S5 to S7 now show the growth phenotype of other selected mutants, one of which includes a start codon mutant.
}

@@

The data regarding greater mutational tolerance in the stalk compared to the head should be paired with data from deep mutational scanning utilizing stalk- and head-specific antibodies as a selection process as this group has previously done with WSN H1 (Doud et al., Nat Comm 2018, Doud et al., PLoS Path 2017) to see if tolerant sites are responsible for viral escape.

\response{We agree that this would be a very interesting experiment to perform.
However, such experiments would be far beyond the scope of our study.
As stated in the Abstract and Significance Statement, the main goal of our study is to perform deep mutational scanning of a recent H3 HA and use the data to better predict H3N2 viral evolution in nature.
A large set of additional experiments comparing stalk- and head-specific antibodies in H3 and H1 HAs is clearly beyond the scope of this goal.
}

Additionally, the authors should directly show the discrepancies between this in vitro selection and the H3 HA's evolution in nature. This group has performed such an analysis in their previous deep mutational scanning work with the H1 HA of WSN (Doud and Bloom, Viruses 2016). Maximum likelihood phylogenetics can be used to estimate the difference between experimental amino acid preference and natural preference. One might expect that this would provide some insight into the nature of the bias introduced through cell culture.

\response{This is a good suggestion and we have now added the results of this analysis in Figure S3.
}

### 2018-06-23 paper/manuscript.tex
more tweaks to reviewer response
finished Jesse revisions

\subsection*{Our measurements are consistent with existing knowledge about HA's evolution and function}
How do the HA amino-acid preferences measured in our experiments relate to the evolution of H3N2 influenza virus in nature?
This question can be addressed by evaluating how well an experimentally informed codon substitution model (ExpCM) using our measurements describes H3N2 evolution compared to standard substitution models~\cite{bloom2017identification,hilton2017phydms}.
Table~\ref{tab:phydms} shows that an ExpCM using the across-replicate average of our measurements greatly outperforms conventional substitution models.
This result indicates that our experiments authentically capture some of the constraints on HA evolution.
A substitution model in which the amino-acid preferences were averaged across all sites (ExpCM, site avg.) performs no better than conventional substitution models, demonstrating that the reason that our measurements are informative is because they capture site-specific constraints.
The relative rate of nonsynonymous to synonymous substitutions (dN/dS or $\omega$) is $\ll$1 for conventional substitution models (Table~\ref{tab:phydms}), since the only way that these models can account for the constraints on HA evolution is via the heuristic use of the $\omega$ parameter to depress the rates of all nonysynonymous substitutions.
However, the relative rate of nonsynonymous to synonymous substitutions after accounting for the amino-acid preferences measured in our experiments ($\omega$ for the ExpCM) is close to one (Table~\ref{tab:phydms}), indicating that most purifying selection against nonsynonymous substitutions is accounted for in the deep mutational scanning.
The ExpCM stringency parameter~\cite{hilton2017phydms} is 2.47 (Table~\ref{tab:phydms}), indicating that natural selection favors the same amino acids as the experiments but with greater stringency.
Throughout the rest of this paper, we use the amino-acid preferences re-scaled~\cite{bloom2017identification,hilton2017phydms} by this stringency parameter.
These re-scaled preferences are shown in Figure~\ref{fig:logoplot}.
Furthermore, Figure~\ref{suppfig:diffprefs_logoplot} shows that when we compare these preferences to amino-acid frequencies from natural human H3N2 HA sequences, at most sites, there is little difference between the amino acids that are preferred in our experiments to their frequencies in nature. \comment{This needs to be improved / changed}

@@

However, there are also some notable differences between the amino-acid preferences measured in our experiments and amino-acid frequencies in natural H3 HA sequences.
Most surprisingly, the start codon does not show a particularly strong preference for methionine (Figure~\ref{fig:logoplot}).
We validated that a virus carrying a mutation at this site from methionine to lysine does in fact reach appreciable titers (Figure~\ref{suppfig:startcodon_validation}), perhaps because of alternative translation-initiation at a downstream or upstream start site as has been described for other HAs~\cite{girard2011upstream}.
Our measurements also suggest mutational tolerance at some other sites that are relatively conserved among natural HAs, such as the N-linked glycosylation motifs near the beginning of HA1 and the transmembrane domain (Figure~\ref{fig:logoplot}).
We validated that viruses with mutations to the glycosylation motifs at sites 22 or 38, or a site in the transmembrane domain, do in fact grow to high titers (Figures~\ref{suppfig:glycan_validation} and \ref{suppfig:TM_validation}, respectively).
The disparity between the relative conservation of these sites in nature and their mutational tolerance in our study could be because cell culture does not fully capture the constraints on HA function on nature, or could be because these sites are not under strong immune pressure and so mutations at them are not positively selected in nature.

@@

\begin{figure}[b!]
\centering
\includegraphics[width=0.95\linewidth]{figs/frequency_trajectory_ex/frequency_trajectory_ex.pdf}
\caption{\label{fig:frequency_trajectory}
{\bf Frequency trajectories of individual mutations and their relation to the experimentally measured effects of these mutations.}
The top panel shows the subset of the full H3N2 HA tree in Figure~\ref{suppfig:tree} from 2004 to 2014.
Circles indicate individual amino-acid mutations, and are colored according to the mutational effect measured in our deep mutational scanning (negative values indicate mutations measured to be deleterious to viral growth).
The Perth/2009 strain is labeled with a star, and nodes in the clade containing the Perth/2009 strain were excluded from our analyses.
The bottom panel shows the frequency trajectory of each mutation, with trajectories colored according to the mutational effects as in the top panel.
It is clear that most mutations that reach high frequency are measured to be relatively favorable in our experiments.
Figure~\ref{suppfig:head_stalk_mutfreq_supp} shows a similar layout but colors mutations by whether they are in HA's head or stalk domain.
}
\end{figure}

@@

Many of the HAs in sequence databases were obtained from virus isolates that were passaged in cell-culture or eggs, which can lead to the acquisition of lab-adaptation mutations that can confound evolutionary analyses~\cite{mcwhite2016sequence}.
To check that our results were not confounded by such lab-adaptation mutations, we repeated our analysis using only HA sequences that were annotated as being derived from viruses that had not been passaged in the lab.
Because sequencing of unpassaged primary isolates has only become commonplace relatively recently, we were only able to perform this analysis for the post-Perth partition of the phylogenetic tree.
Figure~\ref{fig:muteffect_maxfreq}A shows that the correlation between our measured mutational effects and the maximum frequency was even stronger for mutations from unpassaged viral isolates ($\rho = 0.24$).

@@

\subsection*{Measurements made on an H1 HA are not useful for understanding the evolution of H3 influenza}
To determine how broadly experimental measurements can be generalized across HAs, we repeated the foregoing analysis of H3N2 mutation frequencies, but using mutational effects measured in our prior deep mutational scanning of the WSN/1933 H1 HA, which is highly diverged from the Perth/2009 H3 HA (the two HAs only have 42\% protein sequence identity).
Figure~\ref{fig:muteffect_maxfreq_WSN} shows that there is no correlation between the H1 experimental measurements and maximum frequency that mutations reach during H3N2 viral evolution.
Therefore, the utility of an experiment for understanding natural evolution degrades as the experimental sequence becomes more diverged from the natural sequences that are being studied.

@@

Full lab notes from these experiments are at \url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript/tree/master/lab_notes}.

@@

Full lab notes for the library preparation are at \url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript/tree/master/lab_notes}.

\subsection*{Analysis of deep sequencing data}
We used the \texttt{dms\_tools2} software package~\cite{bloom2015software} (\url{https://github.com/jbloomlab/dms_tools2}, version \texttt{2.2.0}) to analyze the deep sequencing data.
The algorithm used to estimate the site-specific amino-acid preferences from the deep sequencing counts is described in~\cite{bloom2015software}.
The amino-acid preferences for each replicate and for the re-scaled, across-replicate average are provided in Dataset~\ref{suppdata:prefsfiles}.
Computer code that performs the entirety of this analysis and shows many detailed plots about read depth and other quality control metrics is available on GitHub at \url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript/blob/master/analysis_code/analysis_notebook.ipynb}.

\subsection*{Phylogenetic model comparison and fitting of a stringency parameter}
For the analysis in Table~\ref{tab:phydms}, we downloaded all full-length H3 HA sequences from the Influenza Virus Resource~\cite{bao2008}, and randomly subsampled two sequences per year.
These sequences were aligned using MAFFT~\cite{katoh2013mafft} and used to infer a phylogenetic tree using RAxML~\cite{stamatakis2006raxml} with a GTRCAT model of nucleotide substitution.
We then used \texttt{phydms}~\cite{hilton2017phydms} (\url{https://github.com/jbloomlab/phydms}, version \texttt{2.2.1}) to fit the substitution models listed in Table~\ref{tab:phydms}.

@@

To determine the effect of virus passaging on the relationship between mutational effects and maximum frequencies in natural populations, we repeated the above analyses using only viruses that were sequenced directly without passaging.
Routine direct sequencing did not begin until the early 2000s \cite{mcwhite2016sequence}.
To construct a tree with a similar number of viruses as the original analysis, we sampled 30 viruses per month between January 1, 2000 and April 1, 2018, producing a tree with 2,374 unpassaged viruses with augur (commit: 6d9f708).
We included the passaged DMS strain, A/Perth/16/2009, in the resulting tree to enable comparison between pre-Perth and post-Perth clades.

@@

\subsection*{Validation of individual point mutants}
To validate the viral growth of Perth/2009 HA point mutants M(-16)K, C52A, C52C, C199(HA2)K, T24F, T40V, and S287A, we used site-directed mutagenesis to introduce the amino-acid mutation into the Perth/2009 HA bidirectional reverse genetics plasmid, and verified the sequence of two clones for each mutant by Sanger sequencing.
We generated these individual mutant viruses carrying GFP in the PB1 segment using a protocol described in \cite{hooper2013mutant}, and the PB2, PA, NP, NA, M, and NS segments of the A/WSN/1933 (H1N1) strain.
To rescue each mutant GFP-carrying virus in duplicate, we transfected a co-culture of $4 \times 10^5$ 293T-CMV-PB1 and $0.5 \times 10^5$ MDCK-SIAT1-CMV-PB1-TMPRSS2 cells with the eight reverse genetics plasmid and the pHAGE2-EF1aInt-TMPRSS2-IRES-mCherry-W plasmid.
Each well received a transfection mixture of 100 $\mu$L DMEM, 3 $\mu$L BioT transfection reagent, and 250 ng of each plasmid.
We changed the media in each well with 2 mL IGM eight hours post-transfection.
At 53 hours post-transfection, transfection supernatants were harvested, clarified by centrifugation at 2000 $\times$ \textit{g} for five minutes, aliquoted, and frozen at -80$^\circ$C.

@@

\clearpage
\subsection*{Supporting Information (SI)}
\clearpage

@@

\begin{suppfigure}[H]
\centerline{\includegraphics[width=0.9\textwidth]{figs/supp_diffprefs/diffprefs.pdf}}
\caption{\label{suppfig:diffprefs_logoplot}
{\bf The differential preferences of the Perth/2009 H3 HA compared to amino-acid frequencies from a sequence alignment of natural human H3.}
The differential preferences estimated by comparing the Perth/2009 H3 HA preferences to amino-acid frequencies from a sequence alignment of natural human H3N2 HA sequences.
For each site, the height of each letter above and below the line indicate how differentially preferred that amino acid is in our experiments compared to in nature.
The overlays show the same information as in Figure~\ref{fig:logoplot} (domain and wildtype amino acid).
The sites are in H3 numbering.
}
\end{suppfigure}

>>>>

\subsection*{Our measurements are consistent with existing knowledge about HA's evolution and function}
How do the HA amino-acid preferences measured in our experiments relate to the evolution of H3N2 influenza virus in nature?
This question can be addressed by evaluating how well an experimentally informed codon substitution model (ExpCM) using our measurements describes H3N2 evolution compared to standard substitution models~\cite{bloom2017identification,hilton2017phydms}.
Table~\ref{tab:phydms} shows that an ExpCM using the across-replicate average of our measurements greatly outperforms conventional substitution models.
This result indicates that our experiments authentically capture some of the constraints on HA evolution.
A substitution model in which the amino-acid preferences were averaged across all sites (ExpCM, site avg.) performs no better than conventional substitution models, demonstrating that the reason that our measurements are informative is because they capture site-specific constraints.
The relative rate of nonsynonymous to synonymous substitutions (dN/dS or $\omega$) is $\ll$1 for conventional substitution models (Table~\ref{tab:phydms}), since the only way that these models can account for the constraints on HA evolution is via the heuristic use of the $\omega$ parameter to depress the rates of all nonysynonymous substitutions.
However, the relative rate of nonsynonymous to synonymous substitutions after accounting for the amino-acid preferences measured in our experiments ($\omega$ for the ExpCM) is close to one (Table~\ref{tab:phydms}), indicating that most purifying selection against nonsynonymous substitutions is accounted for in the deep mutational scanning.
The ExpCM stringency parameter~\cite{hilton2017phydms} is 2.47 (Table~\ref{tab:phydms}), indicating that natural selection favors the same amino acids as the experiments but with greater stringency.
Throughout the rest of this paper, we use the amino-acid preferences re-scaled~\cite{bloom2017identification,hilton2017phydms} by this stringency parameter.
These re-scaled preferences are shown in Figure~\ref{fig:logoplot}.

@@

However, there are also some differences between the amino-acid preferences measured in our experiments and amino-acid frequencies in natural H3 HA sequences (Figure~\ref{suppfig:diffprefs_logoplot}).
Most surprisingly, the start codon does not show a particularly strong preference for methionine (Figure~\ref{fig:logoplot}).
We validated that a virus carrying a mutation at this site from methionine to lysine does in fact reach appreciable titers (Figure~\ref{suppfig:startcodon_validation}), perhaps because of alternative translation-initiation at a downstream or upstream start site as has been described for other HAs~\cite{girard2011upstream}.
Our measurements also suggest mutational tolerance at some other sites that are relatively conserved among natural HAs, such as the N-linked glycosylation motifs near the beginning of HA1 and the transmembrane domain (Figure~\ref{fig:logoplot}).
We validated that viruses with mutations to the glycosylation motifs at sites 22 or 38, or a site in the transmembrane domain, do in fact grow to high titers (Figures~\ref{suppfig:glycan_validation} and \ref{suppfig:TM_validation}, respectively).
The disparity between the relative conservation of these sites in nature and their mutational tolerance in our study could be because cell culture does not fully capture the constraints on HA function on nature, or could be because these sites are not under strong immune pressure and so mutations to them are not positively selected in nature.

@@

\begin{figure}[t!]
\centering
\includegraphics[width=0.95\linewidth]{figs/frequency_trajectory_ex/frequency_trajectory_ex.pdf}
\caption{\label{fig:frequency_trajectory}
{\bf Frequency trajectories of individual mutations and their relation to the experimentally measured effects of these mutations.}
The top panel shows the subset of the full H3N2 HA tree in Figure~\ref{suppfig:tree} from 2004 to 2014.
Circles indicate individual amino-acid mutations, and are colored according to the mutational effect measured in our deep mutational scanning (negative values indicate mutations measured to be deleterious to viral growth).
The Perth/2009 strain is labeled with a star, and nodes in the clade containing the Perth/2009 strain were excluded from our analyses.
The bottom panel shows the frequency trajectory of each mutation, with trajectories colored according to the mutational effects as in the top panel.
It is clear that most mutations that reach high frequency are measured to be relatively favorable in our experiments.
Figure~\ref{suppfig:head_stalk_mutfreq_supp} shows a similar layout but colors mutations by whether they are in HA's head or stalk domain.
}
\end{figure}

@@

Many of the HAs in sequence databases are from viral isolates that were passaged in cell-culture or eggs, which can cause lab-adaptation mutations that confound evolutionary analyses~\cite{mcwhite2016sequence}.
To check that our results were robust to such lab-adaptation mutations, we repeated our analysis using only HA sequences derived from viruses that had not been passaged in the lab.
Because sequencing of unpassaged primary isolates has only recently become commonplace, we could only perform this analysis for the post-Perth partition of the phylogenetic tree.
Figure~\ref{fig:muteffect_maxfreq}A shows that the correlation between our measured mutational effects and the maximum frequency was even stronger for mutations from unpassaged viral isolates ($\rho = 0.24$).

@@

\subsection*{Measurements made on an H1 HA are not useful for understanding the evolution of H3 influenza}
To determine how broadly experimental measurements can be generalized across HAs, we repeated the foregoing analysis of H3N2 mutation frequencies, but using mutational effects measured in our prior deep mutational scanning of the WSN/1933 H1 HA~\cite{doud2016accurate} (Figure~\ref{suppfig:WSNprefs_logoplot}), which is highly diverged from the Perth/2009 H3 HA (the two HAs only have 42\% protein sequence identity).
Figure~\ref{fig:muteffect_maxfreq_WSN} shows that there is no correlation between the H1 experimental measurements and maximum frequency that mutations reach during H3N2 viral evolution.
Therefore, the utility of an experiment for understanding natural evolution degrades as the experimental sequence becomes more diverged from the natural sequences that are being studied.

@@

@@

\subsection*{Analysis of deep sequencing data}
We used the \texttt{dms\_tools2} software package~\cite{bloom2015software} (\url{https://github.com/jbloomlab/dms_tools2}, version \texttt{2.2.5}) to analyze the deep sequencing data.
The algorithm used to estimate the site-specific amino-acid preferences from the deep sequencing counts is described in~\cite{bloom2015software}.
The amino-acid preferences for each replicate and for the re-scaled, across-replicate average are provided in Dataset~\ref{suppdata:prefsfiles}.
Computer code that performs the entirety of this analysis and shows many detailed plots about read depth and other quality control metrics is available on GitHub at \url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript/blob/master/analysis_code/analysis_notebook.ipynb}.

\subsection*{Phylogenetic model comparison and fitting of a stringency parameter}
For the analysis in Table~\ref{tab:phydms}, we downloaded all full-length H3 HA sequences from the Influenza Virus Resource~\cite{bao2008}, and randomly subsampled two sequences per year.
These sequences were aligned using MAFFT~\cite{katoh2013mafft} and used to infer a phylogenetic tree using RAxML~\cite{stamatakis2006raxml} with a GTRCAT model of nucleotide substitution.
We then used \texttt{phydms}~\cite{hilton2017phydms} (\url{https://github.com/jbloomlab/phydms}, version \texttt{2.2.2}) to fit the substitution models listed in Table~\ref{tab:phydms}.

@@

We also repeated the above analyses using only viruses that were sequenced directly without passaging.
Routine direct sequencing did not begin until the early 2000s \cite{mcwhite2016sequence}.
To construct a tree with a similar number of viruses as the original analysis, we sampled 30 viruses per month between January 1, 2000 and April 1, 2018, producing a tree with 2,374 unpassaged viruses with augur (commit: 6d9f708).
We included the passaged DMS strain, A/Perth/16/2009, in the resulting tree to enable comparison between pre-Perth and post-Perth clades.

@@

\subsection*{Validation of individual point mutants}
To validate the viral growth of Perth/2009 HA point mutants M(-16)K, C52A, C52C, C199(HA2)K, T24F, T40V, and S287A, we used site-directed mutagenesis to introduce the amino-acid mutation into the Perth/2009 HA bidirectional reverse genetics plasmid, and verified the sequence of two clones for each mutant by Sanger sequencing.
We generated these individual mutant viruses carrying GFP in the PB1 segment using a protocol described in \cite{bloom2010permissive,hooper2013mutant}, and the PB2, PA, NP, NA, M, and NS segments of the A/WSN/1933 (H1N1) strain.
To rescue each mutant GFP-carrying virus in duplicate, we transfected a co-culture of $4 \times 10^5$ 293T-CMV-PB1 and $0.5 \times 10^5$ MDCK-SIAT1-CMV-PB1-TMPRSS2 cells with the eight reverse genetics plasmid and the pHAGE2-EF1aInt-TMPRSS2-IRES-mCherry-W plasmid.
Each well received a transfection mixture of 100 $\mu$L DMEM, 3 $\mu$L BioT transfection reagent, and 250 ng of each plasmid.
We changed the media in each well with 2 mL IGM eight hours post-transfection.
At 53 hours post-transfection, transfection supernatants were harvested, clarified by centrifugation at 2000 $\times$ \textit{g} for five minutes, aliquoted, and frozen at -80$^\circ$C.

@@

\clearpage
\section*{Supporting Information (SI)}
\FloatBarrier
\pagenumbering{arabic}% resets `page` counter to 1
\renewcommand*{\thepage}{S\arabic{page}}

@@

\begin{suppfigure}[H]
\centerline{\includegraphics[width=0.9\textwidth]{figs/supp_diffprefs/diffprefs.pdf}}
\caption{\label{suppfig:diffprefs_logoplot}
{\bf Sites where there are strong differences between our experimental measurements and the amino-acid frequencies among natural HA sequences.}
We calculated the distance between our H3 measurements and the alignment frequencies using the same approach as in Figure~\ref{fig:distance_distribution}, but using the alignment frequencies in place of the H1 preferences.
For each site, the height of each letter above or below the line indicate how much more or less preferred that amino acid is in our experiments compared to its frequency nature.
The overlays show the same information as in Figure~\ref{fig:logoplot} (domain and wildtype amino acid).
The sites are in H3 numbering.
The HA alignment used to calculate the natural frequencies is the same one used in Table~\ref{tab:phydms}.
\comment{Juhye: the notebook that creates this has a bug, as it never creates the \texttt{H3\_alignment\_frequences\_copy.csv}. 
This must be something you did by hand but then didn't include the code for in the notebook: fix this!
Also, it might be nice if you added a bit more explanation to this section of the Jupyter notebook.
}
}
\end{suppfigure}
